Mixing time estimation in reversible Markov chains from a single sample path

### 2015-06-01 empirical.tex
adding and resolving todos, as per our discussion.
bib file improvements
generalized inverse clean up, both in appendix and in main text
ideas for future work in conclusion
empirical results: comment on O(.) notation
minor tweak to intro

As discussed in \cref{sec:rates}, it is possible to form a confidence
interval for $\gap$ based on the eigenvalues of an estimated
transition probability matrix by appealing to the
Ostrowski-Elsner theorem.
However, as explained earlier, this would lead to a slow
$O(n^{-1/(2d)})$ rate.
We avoid this slow rate by using an estimate of the symmetric matrix
$\vL$, and employing a stronger perturbation result (namely Weyl's
inequality) available for symmetric matrices.

@@

We adopt a different strategy for estimating $\vpi$, which leads to
our construction of empirical confidence intervals, detailed in
\cref{alg:empest}.
We form the matrix $\wh\vP$ using smoothed frequency estimates of
$\vP$ (Step~\ref{step:P}), then 
compute the so-called group inverse $\giAh$ of $\wh\vA = I - \wh\vP$,
followed by finding the unique stationary
distribution $\hat\vpi$ of $\wh\vP$ (Step~\ref{step:pi}).
The group inverse $\giAh$ of $\wh\vA$ is uniquely defined and if $\wh\vP$ defines an ergodic chain
(which is the case here due to the use of the smoothed estimates),
$\giAh$ can be computed at the cost of inverting an $(d-1)\times (d-1)$ matrix
(cf. Theorem 5.2 in \cite{meyer1975role}).
Further, once given $\giAh$, the unique stationary distribution of $\wh\vP$ can be 
read out from the last row of $\giAh$ (Theorem 5.3 of  \cite{meyer1975role}).
The group inverse will also be used later to compute the sensitivity of $\vpi$
based on a result from \cite{cho2001comparison}.
Based on $\hat\vpi$ and $\wh\vP$, we construct the plug-in estimate
$\wh\vL$ of $\vL$, and use the eigenvalues of its symmetrization to
form the estimate $\hatgap$ of the spectral gap (Steps~\ref{step:eig}
and~\ref{step:gap}).
In the remaining steps, 
we use perturbation analyses to relate $\hat\vpi$ and $\vpi$, viewing
$\vP$ as the perturbation of $\wh\vP$; and also to relate $\hatgap$
and $\gap$, viewing $\vL$ as a perturbation of $\Sym(\wh\vL)$.
Both analyses give error bounds entirely in terms of observable
quantities, tracing back to empirical error bounds for the smoothed
frequency estimates of $\vP$.
Note that the most expensive step in the algorithm is the computation of the group inverse,
which, as noted reduces to matrix inversion.
Thus, with a standard implementation of matrix inversion, the algorithm's complexity becomes 
$O( n + d^3)$, while the space complexity is $O(d^2)$.
\todoc{Implement and test?}

@@

The main result of this section is the following result:
\todoc{Conditions?}
\todod{I think they are fine now. Ergodicity is defined in section 3
now, and we discuss why it already implies the $\pimin>0$ and
$\gap>0$.}
\begin{theorem}
  \label{thm:empirical}
  Suppose \cref{alg:empest} is given as input a sample path of length
  $n$ from an ergodic and reversible Markov chain and confidence
  parameter $\delta \in (0,1)$.
  Let $\gap>0$ denote the spectral gap, $\vpi$ the unique stationary
  distribution, and $\pimin>0$ the minimum stationary probability.
  Then, on an event of probability at least $1-\delta$,
  \[
    \pi_i \in [\hat\pi_i-\hat{b},\hat\pi_i+\hat{b}]
    \quad \text{for all $i \in [d]$} ,
    \qquad\text{and}\qquad
    \gap \in [\hatgap-\hat{w},\hatgap+\hat{w}]
    .
  \]
  Moreover, $\hat{b}$ and $\hat{w}$ almost surely satisfy (as $n \to
  \infty$)
  \[
    \hat{b}
    =
    O\Parens{
      \max_{(i,j) \in [d]^2}
      \kappa
      \sqrt{\frac{P_{i,j}\log(d/\delta)}{\pi_i n}}
    }
    ,
    \quad
    \hat{w}
    =
     O\Parens{ 
     \frac{\kappa}{\pimin} \sqrt{\frac{\ln(d/\delta)}{\pimin n}} + 
      \sqrt{ \frac{d\ln(d/\delta)}{ \pimin n}}
    }
    ,
  \]
  where $\kappa$ is analogous to the $\hat{\kappa}$ computed
  by~\cref{alg:empest} with $\giAh$ replaced by the group inverse
  $\giA$ of $\vA := \vI - \vP$.
\end{theorem}
The proof of \cref{thm:empirical} is given in \cref{app:empirical}.
As mentioned above, the obstacle encountered in \cref{thm:err} is
avoided by exploiting the Markov property.
We establish fully observable upper and lower bounds on the entries of
$\vP$ that converge at a $\sqrt{n}$ rate using standard martingale
tail inequalities; this justifies the validity of the bounds from
Step~\ref{step:P-bound}.
Properties of the group inverse~\citep{meyer1975role,cho2001comparison} and eigenvalue
perturbation theory~\citep{stewart1990matrix} are used to validate the
empirical bounds on $\pi_i$ and $\gap$ developed in the remaining
steps of the algorithm.

>>>>

As discussed in \cref{sec:rates}, it is possible to form a confidence
interval for $\gap$ based on the eigenvalues of an estimated
transition probability matrix by appealing to the
Ostrowski-Elsner theorem.
However, as explained earlier, this would lead to a slow
$O(n^{-1/(2d)})$ rate.
We avoid this slow rate by using an estimate of the symmetric matrix
$\vL$, so that we can use a stronger perturbation result (namely Weyl's
inequality, as in the proof of \cref{thm:err}) available for symmetric matrices.

@@

We adopt a different strategy for estimating $\vpi$, which leads to
our construction of empirical confidence intervals, detailed in
\cref{alg:empest}.
We form the matrix $\wh\vP$ using smoothed frequency estimates of
$\vP$ (Step~\ref{step:P}), then compute the so-called group inverse
$\giAh$ of $\wh\vA = I - \wh\vP$ (Step~\ref{step:gi}), followed by
finding the unique stationary distribution $\hat\vpi$ of $\wh\vP$
(Step~\ref{step:pi}), this way decoupling the bound on the accuracy of $\hat\vpi$
from the mixing time.
The group inverse $\giAh$ of $\wh\vA$ is uniquely defined;
 and if
$\wh\vP$ defines an ergodic chain (which is the case here due to the
use of the smoothed estimates), $\giAh$ can be computed at the cost of
inverting an $(d-1)\times (d-1)$ matrix~\citep[Theorem
5.2]{meyer1975role}.%
\footnote{
\label{ftnote:group-inverse}
The group inverse of a square matrix $\vA$, a special case of the {\em Drazin inverse},
is the unique matrix $\vA^\#$ satisfying
$ \vA\vA^\#\vA = \vA$,
$\vA^\#\vA\vA^\# = \vA^\#$ and
$\vA^\#\vA = \vA\vA^\#$.
%It is efficiently computable,
%and for $\vA=\vI-\vP$, where $\vP$ is ergodic with stationary distribution $\vpi$,
%has the
%following simple
%form: $\vA^\#=(\vI-\vP+\v1\vpi^\t)^{-1}-\v1\vpi^\t$
%\citep{seneta1993sensitivity,meyer1975role}.
}
Further, once given $\giAh$, the unique stationary distribution
$\hat\vpi$ of $\wh\vP$ can be read out from the last row of
$\giAh$~\citep[Theorem 5.3]{meyer1975role}.
The group inverse will also be used later to compute the sensitivity
of $\vpi$ based on a result of Cho and Meyer \cite{cho2001comparison}.
Based on $\hat\vpi$ and $\wh\vP$, we construct the plug-in estimate
$\wh\vL$ of $\vL$, and use the eigenvalues of its symmetrization to
form the estimate $\hatgap$ of the spectral gap (Steps~\ref{step:eig}
and~\ref{step:gap}).
In the remaining steps, 
we use perturbation analyses to relate $\hat\vpi$ and $\vpi$, viewing
$\vP$ as the perturbation of $\wh\vP$; and also to relate $\hatgap$
and $\gap$, viewing $\vL$ as a perturbation of $\Sym(\wh\vL)$.
Both analyses give error bounds entirely in terms of observable
quantities, tracing back to empirical error bounds for the smoothed
frequency estimates of $\vP$.
Note that the most expensive step in the algorithm is the computation of the group inverse,
which, as noted reduces to matrix inversion.
Thus, with a standard implementation of matrix inversion, the algorithm's complexity becomes 
$O( n + d^3)$, while its space complexity is $O(d^2)$.
\todoc{Implement and test?}

@@

The main result of this section is the following result:
\begin{theorem}
  \label{thm:empirical}
  Suppose \cref{alg:empest} is given as input a sample path of length
  $n$ from an ergodic and reversible Markov chain and confidence
  parameter $\delta \in (0,1)$.
  Let $\gap>0$ denote the spectral gap, $\vpi$ the unique stationary
  distribution, and $\pimin>0$ the minimum stationary probability.
  Then, on an event of probability at least $1-\delta$,
  \[
    \pi_i \in [\hat\pi_i-\hat{b},\hat\pi_i+\hat{b}]
    \quad \text{for all $i \in [d]$} ,
    \qquad\text{and}\qquad
    \gap \in [\hatgap-\hat{w},\hatgap+\hat{w}]
    .
  \]
  Moreover, $\hat{b}$ and $\hat{w}$ almost surely satisfy (as $n \to
  \infty$)
  \[
    \hat{b}
    =
    O\Parens{
      \max_{(i,j) \in [d]^2}
      \kappa
      \sqrt{\frac{P_{i,j}\log(d/\delta)}{\pi_i n}}
    }
    ,
    \quad
    \hat{w}
    =
     O\Parens{ 
     \frac{\kappa}{\pimin} \sqrt{\frac{\ln(d/\delta)}{\pimin n}} + 
      \sqrt{ \frac{d\ln(d/\delta)}{ \pimin n}}
    }
    ,
  \]
  where $\kappa$ is analogous to the $\hat{\kappa}$ computed
  by~\cref{alg:empest} with $\giAh$ replaced by the group inverse
  $\giA$ of $\vA := \vI - \vP$.
  \todoc{
    Precisely,
    what we mean by the $O(\cdot)$ notation is as follows:
  for a random sequence of nonnegative numbers $(X_n)_n$
  and another sequence of nonrandom positive numbers $(a_n^\eps)_n$ 
  indexed by $\eps$ we
  say that $X_n = O(a_n^\eps)$ holds almost surely as $n\to \infty$
  if for some universal constant $C>0$, 
  $\limsup_{n\to \infty} X_n/a_n^\eps \le C$ holds almost surely, regardless of
  the choice of $\eps$. I think this should be made clear.
  }
\end{theorem}
The proof of \cref{thm:empirical} is given in \cref{app:empirical}.
As mentioned above, the obstacle encountered in \cref{thm:err} is
avoided by exploiting the Markov property.
We establish fully observable upper and lower bounds on the entries of
$\vP$ that converge at a $\sqrt{n}$ rate using standard martingale
tail inequalities; this justifies the validity of the bounds from
Step~\ref{step:P-bound}.
Properties of the group inverse~\citep{meyer1975role,cho2001comparison} and eigenvalue
perturbation theory~\citep{stewart1990matrix} are used to validate the
empirical bounds on $\pi_i$ and $\gap$ developed in the remaining
steps of the algorithm.

### 2015-06-01 intro-nips.tex
adding and resolving todos, as per our discussion.
bib file improvements
generalized inverse clean up, both in appendix and in main text
ideas for future work in conclusion
empirical results: comment on O(.) notation
minor tweak to intro

In this paper we study the construction of fully empirical bounds on the mixing time of Markov chains based on a single sample path.
Let $(X_t)_{t=1,2,\dotsc}$ be an irreducible, aperiodic time-homogeneous Markov chain on a finite state space $[d]$ of $d$ states numbered by positive integers with transition matrix $\vP$. Under this assumption, the chain converges to its unique stationary \todoc{Maybe steady state is better?}
distribution $\vpi = (\pi_i)_{i=1}^d$ regardless of the distribution $\vq$ of $X_1$: 
$\lim_{t\to\infty} \Pr_{\vq}\Parens{X_t = i} = \lim_{t\to\infty} (\vq \vP^t)_i = \pi_i$, $1\le i \le d$.
The mixing time $\tmix$ of the Markov chain is the time that it takes for the chain to reach its stationary distribution, say,
\[
  \tmix
  :=
  \min\Braces{
    t \in \bbN :
    \sup_{\vq}
    \max_{A\subset [d]} |\textstyle\Pr_{\vq}\Parens{ X_t \in A } - \vpi(A)|
%    \Norm{
%      \vq \vP^t - \vpi
%    }_{\tv} 
    \leq 1/4
  }\,.
\]
Here $\vpi(A) = \sum_{i\in A} \pi_i$ is the total probability assigned
to set $A$ by $\vpi$, and the supremum is over
all possible initial distributions $\vq$.
The problem studied in the paper is the construction of a confidence
set $C_n = C_n(X_1,X_2,\dotsc,X_n,\delta) \subset [0,\infty]$ based on
the knowledge of $X_1,X_2,\dotsc,X_n$ only that captures the mixing
time $\tmix$ with a pre-specified probability $1-\delta \in (0,1)$. 
In the remainder of the introduction, we describe the background of why this problem might be interesting, followed by a 
brief description of our main results.
\todod{``might be interesting'' sounds weak.}

>>>>

In this paper we study the construction of fully empirical bounds on the mixing time of Markov chains based on a single sample path.
Let $(X_t)_{t=1,2,\dotsc}$ be an irreducible, aperiodic time-homogeneous Markov chain on a finite state space $[d]$ of $d$ states numbered by positive integers with transition matrix $\vP$. Under this assumption, the chain converges to its unique stationary \todoc{Maybe steady state is better?}
distribution $\vpi = (\pi_i)_{i=1}^d$ regardless of the distribution $\vq$ of $X_1$: 
$\lim_{t\to\infty} \Pr_{\vq}\Parens{X_t = i} = \lim_{t\to\infty} (\vq \vP^t)_i = \pi_i$, $1\le i \le d$.
The mixing time $\tmix$ of the Markov chain is the time that it takes for the chain to reach its stationary distribution, say,
\[
  \tmix
  :=
  \min\Braces{
    t \in \bbN :
    \sup_{\vq}
    \max_{A\subset [d]} |\textstyle\Pr_{\vq}\Parens{ X_t \in A } - \vpi(A)|
%    \Norm{
%      \vq \vP^t - \vpi
%    }_{\tv} 
    \leq 1/4
  }\,.
\]
Here $\vpi(A) = \sum_{i\in A} \pi_i$ is the total probability assigned
to set $A$ by $\vpi$, and the supremum is over
all possible initial distributions $\vq$.
The problem studied in the paper is the construction of a confidence
set $C_n = C_n(X_1,X_2,\dotsc,X_n,\delta) \subset [0,\infty]$ based on
the knowledge of $X_1,X_2,\dotsc,X_n$ only that captures the mixing
time $\tmix$ with a pre-specified probability $1-\delta \in (0,1)$. 
In the remainder of the introduction, we describe our motivation of studying this problem, followed by a 
brief description of our main results.
\todod{``might be interesting'' sounds weak.} \todoc{better now?}

### 2015-06-01 rates.tex
adding and resolving todos, as per our discussion.
bib file improvements
generalized inverse clean up, both in appendix and in main text
ideas for future work in conclusion
empirical results: comment on O(.) notation
minor tweak to intro

Define the random matrix $\wh\vM \in [0,1]^{d \times d}$ and random
vector $\hat\vpi \in \Delta^{d-1}$ by
\begin{align*}
  \wh{M}_{i,j}
  & := \frac{|\{ t \in [n-1] : (X_t,X_{t+1}) = (i,j) \}|}{n-1}
  , \quad i,j \in [d]\,,
  \\
  \hat{\pi}_i
  & := \frac{|\{ t \in [n] : X_t = i \}|}{n}
  , \quad i \in [d]
  .
\end{align*}
Furthermore, define
\[
  \Sym(\wh\vL) := \frac12 \parens{ \wh\vL + \wh\vL^\t }
\]
to be the symmetrized version of the (possibly non-symmetric) matrix
\[
  \wh\vL := \Diag(\hat\vpi)^{-1/2} \wh\vM \Diag(\hat\vpi)^{-1/2}
  .
\]
Let $\hat\lambda_1 \geq \hat\lambda_2 \geq \dotsb \geq \hat\lambda_d$
be the eigenvalues of $\Sym(\wh\vL)$.
Our estimator of the minimum stationary probability $\pimin$ is
\[
  \hatpimin := \min_{i \in [d]} \hat\pi_i ,
\]
and our estimator of the spectral gap $\gap$ is
\[
  \hatgap := 1 - \max\{ \hat\lambda_2, |\hat\lambda_d| \} .
\]

These estimators have the following accuracy guarantees. 
\begin{theorem}
  \label{thm:err}
  There exists an absolute constant $C>0$ such that the following
  holds.
  Assume the estimators $\hatpimin$ and $\hatgap$ described above are
  formed from a sample path of length $n$ from an ergodic and
  reversible Markov chain.
  Let $\gap>0$ denote the spectral gap and $\pimin>0$ the minimum
  stationary probability.
  For any $\delta \in (0,1)$, with probability at least $1-\delta$,
  \begin{equation}\label{eq:piminbound}
    \Abs{\hatpimin-\pimin}
    \le
    C \,
    \Parens{
      \sqrt{\frac{\pimin\log\frac{d}{\pimin\delta}}{\gap n}}
      +
      \frac{\log\frac{d}{\pimin\delta}}{\gap n}
    }
  \end{equation}
  and
  \begin{equation}\label{eq:gapbound}
    \Abs{\hatgap-\gap}
    \leq
    C \,
    \Parens{
      \sqrt{\frac{\log\frac{d}{\delta}\cdot\log\frac{n}{\pimin\delta}}{\pimin\gap n}}
%      + \frac{\log\frac{d}{\delta}\cdot\log\frac{n}{\pimin\delta}}{\pimin\gap n}  
      + \frac{\log\frac{1}{\gap}}{\gap n}  
    }
    \,.
  \end{equation}
\end{theorem}
\todoc{In \eqref{eq:gapbound}, we should be able to merge the second term into the first, right?}
\todod{Oh, I see what you mean---yes, that's right.}

@@

Note that because the eigenvalues of $\vL$ are the same as that of the
transition probability matrix $\vP$, 
we could have instead opted to
estimate $\vP$, say, using simple frequency estimates obtained from
the sample path, and then computing the second largest eigenvalue of
this empirical estimate $\wh\vP$.
The difficulty with this approach is that $\vP$ lacks the structure
required by certain strong eigenvalue perturbation results.
One could instead invoke the Ostrowski-Elsner theorem
\citep[cf.~Theorem 1.4 on Page 170 of][]{stewart1990matrix}, which
bounds the \emph{matching distance} between the eigenvalues of a
matrix $\vA$ and its perturbation $\vA+\vE$ by $O(\norm{\vE}^{1/d})$.
%\footnote{Actually, this result is not even readily applicable: 
%the Haussdorff distance is actually an optimistic estimate of the deviation between
%$\gap$ and its estimate.}
Since $\norm{\wh\vP-\vP}$ is expected to be of size $O(n^{-1/2})$,
this approach will give a confidence interval for $\gap$ whose width
shrinks at a rate of $O(n^{-1/(2d)})$---an exponential slow-down
compared to the rate from \cref{thm:err}.
As demonstrated through an example from \citet{stewart1990matrix}, the
dependence on the $d$-th root of the norm of the perturbation cannot
be avoided in general.
Our approach based on estimating a symmetric matrix affords us the use
of perturbation results that exploit more structure.
%(In \cref{sec:empirical}, we develop another estimator of
%$\vL$---in order to exploit symmetry---that is based on estimating
%$\vP$ for reasons discussed below.)

>>>>

Define the random matrix $\wh\vM \in [0,1]^{d \times d}$ and random
vector $\hat\vpi \in \Delta^{d-1}$ by
\begin{align*}
  \wh{M}_{i,j}
  & := \frac{|\{ t \in [n-1] : (X_t,X_{t+1}) = (i,j) \}|}{n-1}
  , \quad i,j \in [d]\,,
  \\
  \hat{\pi}_i
  & := \frac{|\{ t \in [n] : X_t = i \}|}{n}
  , \quad i \in [d]
  \,.
\end{align*}
Furthermore, define
\[
  \Sym(\wh\vL) := \frac12 \parens{ \wh\vL + \wh\vL^\t }
\]
to be the symmetrized version of the (possibly non-symmetric) matrix
\[
  \wh\vL := \Diag(\hat\vpi)^{-1/2} \wh\vM \Diag(\hat\vpi)^{-1/2}
  .
\]
Let $\hat\lambda_1 \geq \hat\lambda_2 \geq \dotsb \geq \hat\lambda_d$
be the eigenvalues of $\Sym(\wh\vL)$.
Our estimator of the minimum stationary probability $\pimin$ is
\[
  \hatpimin := \min_{i \in [d]} \hat\pi_i ,
\]
and our estimator of the spectral gap $\gap$ is
\[
  \hatgap := 1 - \max\{ \hat\lambda_2, |\hat\lambda_d| \} .
\]

These estimators have the following accuracy guarantees:
\begin{theorem}
  \label{thm:err}
  There exists an absolute constant $C>0$ such that the following
  holds.
  Assume the estimators $\hatpimin$ and $\hatgap$ described above are
  formed from a sample path of length $n$ from an ergodic and
  reversible Markov chain.
  Let $\gap>0$ denote the spectral gap and $\pimin>0$ the minimum
  stationary probability.
  For any $\delta \in (0,1)$, with probability at least $1-\delta$,
  \begin{equation}\label{eq:piminbound}
    \Abs{\hatpimin-\pimin}
    \le
    C \,
    \Parens{
      \sqrt{\frac{\pimin\log\frac{d}{\pimin\delta}}{\gap n}}
      +
      \frac{\log\frac{d}{\pimin\delta}}{\gap n}
    }
  \end{equation}
  and
  \begin{equation}\label{eq:gapbound}
    \Abs{\hatgap-\gap}
    \leq
    C \,
    \Parens{
      \sqrt{\frac{\log\frac{d}{\delta}\cdot\log\frac{n}{\pimin\delta}}{\pimin\gap n}}
%      + \frac{\log\frac{d}{\delta}\cdot\log\frac{n}{\pimin\delta}}{\pimin\gap n}  
      + \frac{\log\frac{1}{\gap}}{\gap n}  
    }
    \,.
  \end{equation}
\end{theorem}

@@

\todoc{Make this paragraph into a remark?}
Note that because the eigenvalues of $\vL$ are the same as that of the
transition probability matrix $\vP$, 
we could have instead opted to
estimate $\vP$, say, using simple frequency estimates obtained from
the sample path, and then computing the second largest eigenvalue of
this empirical estimate $\wh\vP$.
The difficulty with this approach is that $\vP$ lacks the structure
required by certain strong eigenvalue perturbation results.
One could instead invoke the Ostrowski-Elsner theorem
\citep[cf.~Theorem 1.4 on Page 170 of][]{stewart1990matrix}, which
bounds the \emph{matching distance} between the eigenvalues of a
matrix $\vA$ and its perturbation $\vA+\vE$ by $O(\norm{\vE}^{1/d})$.
%\footnote{Actually, this result is not even readily applicable: 
%the Haussdorff distance is actually an optimistic estimate of the deviation between
%$\gap$ and its estimate.}
Since $\norm{\wh\vP-\vP}$ is expected to be of size $O(n^{-1/2})$,
this approach will give a confidence interval for $\gap$ whose width
shrinks at a rate of $O(n^{-1/(2d)})$---an exponential slow-down
compared to the rate from \cref{thm:err}.
As demonstrated through an example from \citet{stewart1990matrix}, the
dependence on the $d$-th root of the norm of the perturbation cannot
be avoided in general.
Our approach based on estimating a symmetric matrix affords us the use
of perturbation results that exploit more structure.
%(In \cref{sec:empirical}, we develop another estimator of
%$\vL$---in order to exploit symmetry---that is based on estimating
%$\vP$ for reasons discussed below.)

### 2015-06-01 setting.tex
adding and resolving todos, as per our discussion.
bib file improvements
generalized inverse clean up, both in appendix and in main text
ideas for future work in conclusion
empirical results: comment on O(.) notation
minor tweak to intro

Define the matrices
\begin{align*}
\vM := \Diag(\vpi) \vP \quad \text{and} \quad
\vL := \Diag(\vpi)^{-1/2} \vM \Diag(\vpi)^{-1/2}\,.
\end{align*}
The $(i,j)$th entry of the matrix $M_{i,j}$ contains the \emph{doublet
probabilities} associated with $\vP$: $M_{i,j} = \pi_i P_{i,j}$ is the
probability of seeing state $i$ followed by state $j$ when the chain
is started from its stationary distribution.
The matrix $\vM$ is symmetric on account of the reversibility of
$\vP$, and hence it follows that $\vL$ is also symmetric.
Further, $\vL = \Diag(\vpi)^{1/2} \vP \Diag(\vpi)^{-1/2}$, hence $\vL$
and $\vP$ are similar and thus their eigenvalue systems are identical.
Ergodicity and reversibility imply that the eigenvalues of $\vL$ are
contained in the interval $\opencloseint{-1,1}$, and that $1$ is an
eigenvalue of $\vL$ with multiplicity $1$~\citep[Lemmas 12.1 and
12.2]{LePeWi08}.
Denote and order the eigenvalues of $\vL$ as
\[
  1 = \lambda_1 > \lambda_2 \geq \dotsb \geq \lambda_d > -1 .
\]
Let $\slem := \max\{ \lambda_2,\, |\lambda_d| \}$, and define the
(absolute) spectral gap to be $\gap := 1-\slem$, which is strictly positive.
%Throughout the paper, both $\pimin$ and $\gap$ are assumed
%to be positive.

@@

The goal is to estimate $\pimin$ and $\gap$ from the length $n$ sample
path $(X_t)_{t \in [n]}$, and also to construct fully empirical
confidence intervals that $\pimin$ and $\gap$ with high probability;
in particular, the construction of the intervals should not depend on
any unobservable quantities, including $\pimin$ and $\gap$ themselves.
It is well-known that the \emph{mixing time} of the Markov chain $\tmix$ (defined in the introduction)
\hide{,
\[
  \tmix
  :=
  \min\Braces{
    t \in \bbN :
    \sup_{\vq \in \Delta^{d-1}}
    \Norm{
      \vq \vP^t - \vpi
    }_{\tv} \leq 1/4
  }
  ,
\]}
can be bounded in terms of $\pimin$ and $\gap$~\citep[Theorems~12.3
and~12.4]{LePeWi08}:
\[
  \Parens{\frac1{\gap} - 1} \ln2
  \ \leq \ \tmix
  \ \leq \ \frac1{\gap} \log \frac4{\pimin}
  .
\]
Moreover, convergence rates for empirical processes on Markov chain
sequences are also often given in terms of mixing coefficients that
can ultimately be bounded in terms of $\pimin$ and
$\gap$~\citep[e.g.,][]{MR1921877,MoRo08,MoRo09}. \todoc{Cite one? Btw, do these really have these results?}
Therefore, valid confidence intervals for $\pimin$ and $\gap$ can be
used to make these rates fully observable.

>>>>

Define the matrices
\begin{align*}
\vM := \Diag(\vpi) \vP \quad \text{and} \quad
\vL := \Diag(\vpi)^{-1/2} \vM \Diag(\vpi)^{-1/2}\,.
\end{align*}
The $(i,j)$th entry of the matrix $M_{i,j}$ contains the \emph{doublet
probabilities} associated with $\vP$: $M_{i,j} = \pi_i P_{i,j}$ is the
probability of seeing state $i$ followed by state $j$ when the chain
is started from its stationary distribution.
The matrix $\vM$ is symmetric on account of the reversibility of
$\vP$, and hence it follows that $\vL$ is also symmetric.
As we will see later, we will strongly exploit this later.  \todoc{Say more? Maybe not..}
%will be advantageous due to the existence of stronger perturbation results
%for symmetric matrices.
Further, $\vL = \Diag(\vpi)^{1/2} \vP \Diag(\vpi)^{-1/2}$, hence $\vL$
and $\vP$ are similar and thus their eigenvalue systems are identical.
Ergodicity and reversibility imply that the eigenvalues of $\vL$ are
contained in the interval $\opencloseint{-1,1}$, and that $1$ is an
eigenvalue of $\vL$ with multiplicity $1$~\citep[Lemmas 12.1 and
12.2]{LePeWi08}.
Denote and order the eigenvalues of $\vL$ as
\[
  1 = \lambda_1 > \lambda_2 \geq \dotsb \geq \lambda_d > -1 .
\]
Let $\slem := \max\{ \lambda_2,\, |\lambda_d| \}$, and define the
(absolute) spectral gap to be $\gap := 1-\slem$, which is strictly positive.
\todoc{Do our results generalize to the spectral gap? Where do we need that we talk about the absolute spectral gap?
What is the difference between the spectral and the absolute spectral gap? Say, in their relationship to $\tmix$?}
%Throughout the paper, both $\pimin$ and $\gap$ are assumed
%to be positive.

@@

The goal is to estimate $\pimin$ and $\gap$ from the length $n$ sample
path $(X_t)_{t \in [n]}$, and also to construct fully empirical
confidence intervals that $\pimin$ and $\gap$ with high probability;
in particular, the construction of the intervals should not depend on
any unobservable quantities, including $\pimin$ and $\gap$ themselves.
It is well-known that the \emph{mixing time} of the Markov chain $\tmix$ (defined in the introduction)
\hide{,
\[
  \tmix
  :=
  \min\Braces{
    t \in \bbN :
    \sup_{\vq \in \Delta^{d-1}}
    \Norm{
      \vq \vP^t - \vpi
    }_{\tv} \leq 1/4
  }
  ,
\]}
can be bounded in terms of $\pimin$ and $\gap$~\citep[Theorems~12.3
and~12.4]{LePeWi08}:
\[
  \Parens{\frac1{\gap} - 1} \ln2
  \ \leq \ \tmix
  \ \leq \ \frac1{\gap} \log \frac4{\pimin}
  .
\]
Moreover, convergence rates for empirical processes on Markov chain
sequences are also often given in terms of mixing coefficients that
can ultimately be bounded in terms of $\pimin$ and
$\gap$, as we will show in the proof of our first result.\todoc{I removed the citations (saving space), referring to the proof.}
%~\citep[e.g.,][]{MR1921877,MoRo08,MoRo09}. 
Therefore, valid confidence intervals for $\pimin$ and $\gap$ can be
used to make these rates fully observable.
